Polynomial Time and Dependent Types

### 2023-11-14 paper/paper-extended.tex
[ paper ] sync short and extended versions
[ paper ] edits

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The systems are presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Bounded-error
  Probabilistic Polynomial (BPP) time. Since these classes rely on
  specific correctness criteria, it is not possible to capture them in
  a simply typed system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

It is not too difficult to see that reduction of linear
$\lambda$-terms always takes a number of steps linearly proprotional
to the size of the term. This is because every $\beta$-redex
substitutes each term into at most one variable, reducing the size of
the term by one each time.

We can increase the expressivity, but not the computational
complexity, of the system by adding datatypes that do not allow
iteration. These can be used for representation, but not for driving
computation. We include the rules here, to show how linearity must be
preserved in these rules, and to foreshadow their dependently typed
counterparts in \autoref{sec:noniter-qtt}. The first type is the
booleans, which are anyway non-recursive and so would not allow
iteration anyway:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cTrue, \cFalse : \BoolTy}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aelhig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT. The full
rules, including all equality rules, are presented in
\autoref{sec:rules}.

@@

\paragraph{Linear and Substructural Dependent Types} We chose QTT as
the particular combination of linear and dependent types for our
systems. Other systems include systems such as those by
\citet{CervesatoP02}, \citet{KrishnaswamiPB15}, and \citet{Vakar14}
which all use a strict separation between linear and non-linear
variables. This strict separation would mean that we could not as
easily move programs from the linear fragment into the types, as in
\autoref{sec:programming-polytime}. Systems that are more like QTT in
that they do not have a strict separation of variables include those
of \citet{MoonEO21}, \citet{ChoudhuryEEW21}, and \citet{Abel18}. These
systems differ from QTT in that they do not include a complete copy of
unrestricted type theory as QTT does in its $\sigma = 0$ fragment,
because they all track usage in types as well as terms, so it is not
clear how to use them for unrestricted reasoning as we do with
QTT. \citet{FuKS22} present a system that is closer to QTT but does
not include a universe type, which we used in \autoref{sec:classes} to
be able to characterise complexity classes as predicates decidable in
restricted complexity.

@@

>>>>

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The systems are presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a polytime sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Bounded-error
  Probabilistic Polynomial (BPP) time. Since these classes rely on
  specific semantic correctness criteria, it is not possible to capture them in
  a simply typed system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness} and
  \autoref{sec:realising-iteration}. Our construction is an extension
  of the amortised complexity realisability constructions of
  \citet{dallago11realisability}. We extend their work to our
  dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of these sections has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the associated
  artefact \cite{atkey2023polydep-artefact}.
\end{enumerate}

@@

It is not too difficult to see that reduction of linear
$\lambda$-terms always takes a number of steps linearly proportional
to the size of the term. This is because every $\beta$-redex
substitutes each term into at most one variable, reducing the size of
the term by one each time.

We can increase the expressivity, but not the computational
complexity, of the system by adding datatypes that do not allow
iteration. These can be used for representation but not for driving
computation. We include the rules here to show how linearity must be
preserved in these rules and to foreshadow their dependently typed
counterparts in \autoref{sec:noniter-qtt}. The first type is the
booleans, which are non-recursive and so would not allow
iteration anyway:
\begin{mathpar}
  \inferrule*
  { }
  {\vdash \cTrue, \cFalse : \BoolTy}

@@

  \inferrule*
  {\Gamma_1 \vdash M : \Diamond \\ \Gamma_2 \vdash N : \tyNat}
  {\Gamma_1, \Gamma_2 \vdash \conSucc(M, N) : \tyNat}
\end{mathpar}
To construct a $\conZero$, we must have a $\Diamond$ to pay for it,
and likewise, to construct a $\conSucc$ we must pay a $\Diamond$. We
can think of $\Diamond$s as an unit of iterable data. Iterability is
``saved up'' in data during construction, and released during
iteration. Diamonds cannot be created by a program itself, for the
same reason that constructors were prohibited in the \emph{Cons-free}
system, but they are released from iterable data during iteration. The
LFPL natural number iterator has the following typing rule:
\begin{displaymath}
  \inferrule*
  {d : \Diamond \vdash M_z : A \\ d : \Diamond, x : A \vdash M_s : A \\ \Gamma \vdash N : \tyNat}
  {\Gamma \vdash \tmRec\,N\,\{\conZero(d) \mapsto M_z; \conSucc(d,x) \mapsto M_s \} : A}
\end{displaymath}
The difference with the \emph{Cons-free} iterator above is that the
$\conZero$ and $\conSucc$ cases now both have an additional binding of
type $\Diamond$. This allows some form nesting of iterations: during
an iteration over the input, the program can accumulate $\Diamond$s to
use for iteration over substructures that are smaller than the current
point in the iteration. A construction, due to \citet{AehligS02},
illustrates how this leads to all polytime computations. As above, we
assume a polynomial $p(n)$ and a step function
$f : \mathit{St} \multimap \mathit{St}$ that needs to be iterated
$p(n)$ times. We construct a linear iterator:
\begin{displaymath}
  \begin{array}{l}
    I_1 : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_1 = \lambda (n, s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d), s);\\
      \conSucc(d,(n,s)) &\mapsto& (\conSucc(d,n),f\,s) \}
    \end{array}
  \end{array}
\end{displaymath}
Note that this iterator returns the natural number input as well as
the new state. LFPL does not allow duplication of iterable inputs, so
we must always reconstruct it if we want to do further
iteration. Addition of polynomials is accomplished by composition of
iterators. To raise the degree, we again use a nesting iterator:
\begin{displaymath}
  \begin{array}{l}
    I_{k+1} : (\tyNat \otimes \mathit{St}) \multimap (\tyNat \otimes \mathit{St}) \\
    I_{k+1} = \lambda (n,s).\,\tmRec\,n\,\{
    \begin{array}[t]{@{}lcl}
      \conZero(d)&\mapsto&(\conZero(d),s); \\
      \conSucc(d,(n,s)) &\mapsto& \Let\,(n,s)=I_k\,(n,s)\,\In\,(\conSucc(d,n), s) \}
    \end{array}
  \end{array}
\end{displaymath}
Unlike in the \emph{Cons-free} system, this iterator does not raise
the degree of the nested iterator directly. Rather, the iterator $I_k$
on the input $n$ performs $\binom{n}{k}$ iterations. As observed by
Aehlig and Schwictenberg, this is sufficient because the binomials
form a basis for the vector space of all polynomials.

@@

In this section, we give an overview of the rules of QTT. The full
rules, including all equality rules, are presented in
\autoref{sec:rules}.
%the online appendix \cite{atkey2023polydep-online}.

@@

\paragraph{Linear and Substructural Dependent Types} We chose QTT as
the particular combination of linear and dependent types for our
systems. Other systems include systems such as those by
\citet{CervesatoP02}, \citet{KrishnaswamiPB15}, and \citet{Vakar14}
which all use a strict separation between linear and non-linear
variables. This strict separation would mean that we could not as
easily move programs from the linear fragment into the types, as in
\autoref{sec:programming-polytime}. Systems that are more like QTT in
that they do not have a strict separation of variables include those
of \citet{MoonEO21}, \citet{ChoudhuryEEW21}, and \citet{Abel23}. These
systems differ from QTT in that they do not include a complete copy of
unrestricted type theory as QTT does in its $\sigma = 0$ fragment,
because they all track usage in types as well as terms, so it is not
clear how to use them for unrestricted reasoning as we do with
QTT. \citet{FuKS22} present a system that is closer to QTT but does
not include a universe type, which we used in \autoref{sec:classes} to
be able to characterise complexity classes as predicates decidable in
restricted complexity.

@@

\section*{Data Availability Statement}

The Agda source files and rendered HTML for this paper is available from Zenodo \cite{atkey2023polydep-artefact}. The source files are also available online at GitHub: \url{https://github.com/bobatkey/qtt-models}, commit SHA1
